Finiteness in Cubical Type Theory

### 2020-01-16 paper.tex
more

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In constructive mathematics, we are often preoccupied with \emph{how} we know
something.
Finiteness is one of the prime examples which demonstrates this phenomenon: to
show that some type is finite, we might provide a bijection with another
finite set, or a surjection from some finite type.
We could alternatively say that a type is finite if any collection of its
elements larger than some bound contains duplicates, or if any stream of its
elements contains duplicates.

Classically, all of these ways to prove finiteness end up proving the same
thing: constructively, however, each defines a fundamentally different
``version'' of finiteness.
In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
variations (and more) have been explored in
\cite{spiwackConstructivelyFinite2010} and
\cite{firsovVariationsNoetherianness2016}.

@@

\subsection{Strong Finiteness Predicates}
\input{figures/finite-classification}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

>>>>

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In constructive mathematics, we are often preoccupied with \emph{how} we know
something, in contrast to the classical focus on the \emph{what}.
Take finiteness, for example.
There are a handful of ways to demonstrate some type is finite: we could provide
a bijection or a surjection from another finite type; we could show that any
collection of its elements larger than some bound contains duplicates, or that
any stream of its elements contain duplicates.

Classically, all of these proofs end up proving the same thing: that our type is
finite.
Constructively, however, all \emph{four} of the statements above imply a
different ``version'' of finiteness.
\emph{How} we show that some type is finite has a significant impact on the
scope of the things we can do with that type.

That is the first thing this paper sets out to investigate: different finiteness
predicates.
The five in particular that we are interested in are organised in
figure~\ref{finite-classification}.
We especially want to examine what separates each predicate: what extra
information do you need, for instance, to go from a proof of Cardinal finiteness
(section~\ref{cardinal}) to manifest Bishop finiteness
(section~\ref{manifest-bishop}).
These predicates aren't just different in proof obligation, either: there are
concrete types which can be shown to satisfy some, but not all, of these
finiteness predicates.

\input{figures/finite-classification}

Next, we will examine the category-theoretic and topos-theoretic interpretations
of each of the predicates.

ways to prove finiteness end up proving the same
thing: constructively, however, each defines a fundamentally different
``version'' of finiteness.
In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
variations (and more) have been explored in
\cite{spiwackConstructivelyFinite2010} and
\cite{firsovVariationsNoetherianness2016}.

@@

\subsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.
